A Recipe for Scavenging - the natural history of a behaviour

### 2016-03-31 natural history of scavenging.tex
abstract and conclusion

\begin{abstract}
 Scavengers existed in the past and they exist now. 
 Often under appreciated. 
 Three main habitat types considered: land, air and sea. 
 Different drivers in these areas. 
 Review looks at these 
\end{abstract}

@@

\subsection*{The Difficulty of Scavenging} %KH: Make that the template for all the next sections
The chief hurdle to scavenging is finding a sufficient quantity of food, the occurence of which is difficult to predict in space and time. Thus, any animal existing as a scavenger must minimise its locomotory costs and maximise its detection capabilities \citep{ruxton2004obligate}.
Once found, the scavenger has to process the carrion and overcome the agents of decay produced by the action of microorganisms on the carcass \citep{ruxton2014fruit}. 
The habitat must also be productive enough to sustain an animal biomass that will eventually produce carcasses.
We can draw on the image of a scavenger moving through its environment, searching for food and trying to process it efficiently as we explore the prevalence of this behaviour through time. 
% The idea of scrounging from predator kills is undermined by studies showing that in the  majority of ecosystems more animals die from disease and starvation than predation \citep{benbow2015introduction}.

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{Summary_figure/Summary_figure.pdf}
\caption{Factors influencing the proportion of scavenging in a vertebrates' diet. Blue lines indicates a reduction in the factor and red lines indicates an increase.}
%TG: Probably needs some more explanations...
\label{Summary_figure}
\end{figure}

@@

% \section*{Aerial Scavengers}
% Vultures represent 
%These birds consist of two convergent groups, from the old and the new world and represent the only example of obligate vertebrate scavengers today.
%The families from which modern vultures arose, the Accipitridae and Cathartidae, appear during the Palaeocene (66 - 56 Mya) \citep{Jetz2012, Jarvis2014}.
%Given their unique position, they have been extensively studied to determine what adaptations they possess that allows them to so flourish in this niche.
%Underscoring their prowess are studies such as \cite{ogada2012effects} who note a three fold increase in carcass persistence times in the absence of vultures.
%As such, we can begin by exploring the adapdations and the environments of vultures to draw comparisons with other scavenging species and \textit{their} environments.

\subsection*{Locomotion}
As endotherms, mammals can sustain long bouts of energetically expensive activity. 
By contrast, modern reptiles are ectothermic, limiting their activity periods.
% a successful reptilian scavenger requires a far different set of adapations.
This is exacerbated by the sprawling gait seen in many lizards which results in Carrier's Constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}.
This constraint manifests itself in aspects such as maximum sustainable speed where an equivalent mammal has a six to seven fold increase \citep{ruben1995evolution}.
To quantify this effect with a simple example we can turn to some allometric relationships relating sustainable travelling speed to body mass.
In the case of mammals and reptiles these are 1.15 $\times$ body mass (kg) \textsuperscript{0.12} and 0.23 $\times$ body mass (kg) \textsuperscript{0.12} respectively \citep{ruxton2004obligate}.
We can insert these into a foraging radius model $\sfrac{\frac{\text{duration} \times \text{speed}}{2}}{1000}$ for a 12 hour foraging day which shows that while a 10 kg reptile can range 6.5 km an equally sized mammal can range nearly 33 km \citep{Enstipp2006Energetics}. 
For a foraging scavenger, this ability translates into a greater area searched for food.

@@

%TG: I think it's missing a transition here: maybe something along "But moving on land as it's limitations..."
But it is in the air that we find scavengers par excellence. 
Flight is a cheaper means of locomotion than walking or running \citep{tucker1975energetic}. 
We know that many extant birds exist as facultative scavengers because storks, raptors and corvids all take substantial quantities of carrion in their diet \citep{kendall2013alternative}.
The advantage of flight can be extended further in larger species that engage in soaring instead of flapping flight, which is even cheaper still (approximately twice the basal metabolic rate) \citep{hedenstrom1993migration,spivey2014analysing}.
The benefits this confers are clear from the information we have on the enormous foraging ranges of many vultures \citep{spiegel2013factors} and seabirds \citep{thaxter2012seabird}.
In vultures we have the best known scavengers on Earth. 
These birds consist of two convergent groups, from the old and the new world where they represent the only example of obligate vertebrate scavengers today.
The families from which modern vultures arose, the Accipitridae and Cathartidae, appear during the Palaeocene \citep[66 - 56 Mya; ][]{Jetz2012, Jarvis2014}.
Yet, avian flight is far older than this and originates in the Late Jurassic (163.5-145 Mya), conincident with the fossils of \textit{Archaeopteryx lithographica} so many of these benefits would have been realised from that point on for carnivorous birds.
And vertebrate flight is much older still where pterosaurs predate bird origins by a considerable margin in the Late Triassic (235-201.3 Mya).
Scavenging in this diverse group has been hypothesied many times before \citep{witton2008reappraisal}.
Certain clades of these animals could reach enormous sizes \citep[e.g. Azhdarchids with wingspans of 11 metres; ][]{witton2010size} and, notably, look to have engaged in soaring flight \citep{witton2010size}.
%TG: this paragraph's a killer! Love it!

@@

\subsection*{Detection}
It would be pointless to have incredible ranging abilites and not have the sensory architecture to benefit from it.
If we came at this from a position of complete ignorance we would predict scavengers to have well-developed senses and indeed this is what we find. 
A simplification of terrestrial, vertebrate scavengers in sensory terms is one of them existing in a two-dimensional plane while foraging for carrion directly.
They can detect carcasses at a range that is defined by the radius of their sensory organs. %, usually the visual and olfactory senses.
As a consequence, they have a much more restricted view of the landscape than do aerial foragers.
Hyenas make up for this in their ability to smell a rotting carcass 4 km away and to hear the vocalisations of conspecifics at a distance of 10 km \citep{mills1989comparative}. 
% Using the approach of \cite{spiegel2013factors} we estimate a spotted hyena %TG: do we need a (\textit{Crocuta crocuta}) here?
%could resolve a 2 metre target at 1 km distance. 
While considering prehistoric habitats \cite{ruxton2004obligate} calculated that ``a 1 tonne mammal or reptile, in an ecosystem yielding carrion at densities similar to the current Serengeti, could have met its energy requirements if it could detect carrion over a distance of the order of 400–500 m".
The senses of many extant (and in all probabilty extinct) carnivores meet this required distance, making scavenging feasbile for terrstrial species \citep{farlow1994speculations,mech2010wolves}.

@@

Aside from sight, three species of vultures within the new world family Cathartidae, (genus \textit{Cathartes}), have well developed olfactory systems \citep{AR:AR22815}.
Among them are the Turkey Vultures (\textit{Cathartes aura}) which were able to locate 90\% of baits set out in a tropical forest \citep{houston1986olfaction}. An atuned sense of smell is obviously useful in detecting decaying carrion from the air.
 %TG: this sentence makes totally sense but starting it by "Clearly" kind of undermines it. To me it sounds like when in a presentation some spawn of Satan puts up a table and starts saying: "As you can see".
% This would be impossible for the visually reliant old world species.

@@

% Depending on the species, a carcass in water either floats or descends to the sea floor \citep{Whitehead415}.
Water tends to be a low-light environment, so visual detection distances are far lower (< 100 m) than they would be in the air.
As such, animals detect resources through chemo- and mechanoreception more so than through vision \citep{ruxton2004energetic}.
This is particularly relevant to extant sharks and aquatic snakes who are deemed as having the most suitable physiology for scavenging.
A hypothesis put forth by \cite{sazima1990necrofagia} argued that chemical gradients in water would allow for a relatively easier detection of carrion by snakes.
This gained some support from \cite{devault2002scavenging}, who found a preponderence of aquatic snake species in their review of this behaviour.
Smell seems to be the primary means of carcass detection in sharks as well. 
\cite{fallows2013white} found that wind speed determined the number of sharks feeding at whale carcasses indicating they were dependent on detecting the odours from the decaying whales.

@@

Interestingly, such comparisons have given insight into the feeding ecology of early hominins who, for instance, had the ability to craft tools for breaking open bones \citep{ARCM:ARCM12084}.
The question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene has been a matter of debate for years.
A recent study investigating potential scavenging opportunities for hominins in Kenya found that, even when discounting bone material, there is a substantial amount of scavengeable meat left on predated remains; sufficient to sustain the requirements of an adult male \textit{Homo erectus} \citep{pobiner2015new}.
In some historical hominin-inhabited areas there were a greater number of felids than hyenids.
This is significant because hyenas are likely to have left far less flesh on a carcass than a felid such as a sabretooth enabling contemperaneous hominins to benefit \citep{pobiner2015new}.
The intelligence, resultant tool-use and cooperative nature of hominins meant they could likely adapt to take on more or less carrion depending on their environment \citep{moleon2014humans}.
%TG: or "could"? This is hypothetical behaviour stuff! Maybe past hyenas where exclusive sushi eaters!
%TG: Not sure about the intelligence thing: the sentence as the same meaning without and will be less anthropocentric sounding.

@@

Scavenging should be particularly attractive to avian predators compared to mammals.
Solitary mammalian predators can kill prey up to the same body mass as themselves and sometimes an order of magnitude heavier (e.g. socially hunting lions \citep{owen2008predator}).
In contrast, birds of prey tend to kill prey smaller than themselves \citep{slagsvold2007prey}.
This is likely due to their need to kill animals that they can fly away with, as well as the risk of injury being higher (which carries a higher mortality risk) for a bird than a mammal.
Scavenging provides a means for birds to exploit species that would otherwise be too big for them to kill.
% Thus, for birds, scavenging means they can exploit species that would otherwise be too big for them to kill.

@@

Perhaps the greatest environmental driver of scavenging tendency is that of temperature. 
The geological record shows the Earth has undergone radical fluctuations in temperature.
This will have had a significant bearing on the availability and persistence of carrion.
To illustrate the point, a 10$^{\circ}$C increase in ambient temperature can double carcass decomposition rates \citep{parmenter2009carrion} and geological evidence indicates that the Mesozoic Earth was at least 6 $^{\circ}$C warmer than now \citep{sellwood2006mesozoic}.
In terms of specific habitats, it has been shown that decomposition is greater in warm and moist areas versus more xeric ones \citep{beasley2015vertebrates}.
Moreover, oceanic productivity and habitat structure are all impacted by climactic conditions.
The impacts these can have on scavengers have been empirically supported e.g.
\cite{beasley2015vertebrates} who point to a series of studies showing how microbes and invertebrates benefit at higher temperatures to the detriment of vertebrate scavengers such that ``above 20$^{\circ}$C vertebrates were able to detect and consume only 19\% of small-mammal carcasses, whereas at temperatures below 18$^{\circ}$C, vertebrates consumed 49\% of carcasses".

%\section*{The Role of Climate} %TG: Use it as a opening at the end of the paper (i.e. conclusion + finishing with this bit)
\section*{Conclusion} 
As is often the case in science, the present provides the key to the past. The animals of today, while often different (sometimes radically so) to their ancestors, allow us to make informed comparisons to extinct species. We have used this technique to give insight into the drivers of scavenging across terrestrial vertebrates through time. 
In common with any other forager be they grazer, browser or predator, scavengers past and present have had to balance their energetic costs with the gains of food. 
% Add bits on climate change/dectection rate change, etc...

>>>>

\begin{abstract} 
% Here, we chart the natural history of scavenging. 
Despite its prevalence, scavenging is a difficult behaviour to observe in modern day carnivores and impossible to study directly in extinct species. 
Yet, there are certain intrinsic and environmental features of a species that push it towards a scavenging lifestyle. Chief among these are low-cost locomotion, high detection distances, effective carcass processing and a carrion-producing habitat. 
We use these four components to highlight the morphologies and environments that would have been conducive to scavenging over geological time by focussing on dominant vertebrate groups of the land, sea and air. 
The result is a document on the natural history of scavenging, the first to our knowledge. 
Our idea of a scale of scavenging can be applied to any species at any time to judge the importance of this behaviour in its diet. 
\end{abstract}

@@

\subsection*{The Difficulty of Scavenging} %KH: Make that the template for all the next sections
The chief hurdle to scavenging is finding a sufficient quantity of food, the occurence of which is difficult to predict in space and time. Thus, any animal existing as a scavenger must minimise its locomotory costs and maximise its detection capabilities \citep{ruxton2004obligate}.
Once found, the scavenger has to process the carrion and overcome the agents of decay produced by the action of microorganisms on the carcass in order to take in a suffienct amount of energy \citep{ruxton2014fruit}. 
The habitat must also be productive enough to sustain an animal biomass that will eventually produce carcasses.
We can draw on the image of a scavenger moving through its environment, searching for food and trying to process it efficiently as we explore the prevalence of this behaviour through time. 
% The idea of scrounging from predator kills is undermined by studies showing that in the  majority of ecosystems more animals die from disease and starvation than predation \citep{benbow2015introduction}.

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{Summary_figure/Summary_figure_Landscape.pdf}
\caption{Factors influencing the proportion of scavenging in a vertebrates' diet. Blue lines indicates a reduction in the factor and red lines indicates an increase.}
%TG: Probably needs some more explanations...
\label{Summary_figure}
\end{figure}

@@

\subsection*{Locomotion}
As endotherms, mammals can sustain long bouts of energetically expensive activity. 
By contrast, modern reptiles are ectothermic, limiting their activity periods.
% a successful reptilian scavenger requires a far different set of adapations.
This is exacerbated by the sprawling gait seen in many lizards which results in Carrier's Constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}.
These constraints manifests themselves in aspects such as maximum sustainable speed where an equivalent mammal has a six to seven fold increase \citep{ruben1995evolution}.
To quantify this effect with a simple example we can turn to some allometric relationships relating sustainable travelling speed to body mass.
In the case of mammals and reptiles these are 1.15 $\times$ body mass (kg) \textsuperscript{0.12} and 0.23 $\times$ body mass (kg) \textsuperscript{0.12} respectively \citep{ruxton2004obligate}.
If we insert these into a foraging radius model $\sfrac{\frac{\text{duration} \times \text{speed}}{2}}{1000}$ for a 12 hour foraging day which shows that while a 10 kg reptile can range 6.5 km an equally sized mammal can range nearly 33 km \citep{Enstipp2006Energetics}. 
For a foraging scavenger, this ability translates into a greater area searched for food.

@@

%TG: I think it's missing a transition here: maybe something along "But moving on land as it's limitations..."
But it is in the air that we find scavengers par excellence. 
Flight is a cheaper means of locomotion than either walking or running \citep{tucker1975energetic}. 
We know that many extant birds exist as facultative scavengers because storks, raptors and corvids all take substantial quantities of carrion in their diet \citep{kendall2013alternative}.
The advantage of flight can be extended further in larger species that engage in soaring instead of flapping flight, which is even cheaper still (approximately twice the basal metabolic rate) \citep{hedenstrom1993migration,spivey2014analysing}.
The benefits this confers are clear from the information we have on the enormous foraging ranges of many vultures \citep{spiegel2013factors} and seabirds \citep{thaxter2012seabird}.
In vultures we have the best known scavengers on Earth. 
These birds consist of two convergent groups, from the old and the new world where they represent the only example of obligate vertebrate scavengers today.
The families from which modern vultures arose, the Accipitridae and Cathartidae, appear during the Palaeocene \citep[66 - 56 Mya; ][]{Jetz2012, Jarvis2014}.
Yet, avian flight is far older than this and originates in the Late Jurassic (163.5-145 Mya), conincident with the fossils of \textit{Archaeopteryx lithographica}, so many of these benefits would have been realised from that point on for carnivorous birds.
And vertebrate flight is much older still where pterosaurs predate bird origins by a considerable margin in the Late Triassic (235-201.3 Mya).
Scavenging in this diverse group has been hypothesied many times before \citep{witton2008reappraisal}.
Certain clades of these animals could reach enormous sizes \citep[e.g. Azhdarchids with wingspans of 11 metres; ][]{witton2010size} and, notably, look to have engaged in soaring flight \citep{witton2010size}.
%TG: this paragraph's a killer! Love it!

@@

\subsection*{Detection}
It would be pointless to have incredible ranging abilites and not have the sensory architecture to benefit from it.
If we came at this from a position of complete ignorance we would predict scavengers to have well-developed senses and indeed, this is what we find. 
A simplification of terrestrial, vertebrate scavengers in sensory terms is one of them existing in a two-dimensional plane while foraging for carrion directly.
They can detect carcasses at a range that is defined by the radius of their sensory organs. %, usually the visual and olfactory senses.
As a consequence, they have a much more restricted view of the landscape than do aerial foragers.
Hyenas make up for this in their ability to smell a rotting carcass 4 km away and to hear the vocalisations of conspecifics at a distance of 10 km \citep{mills1989comparative}. 
% Using the approach of \cite{spiegel2013factors} we estimate a spotted hyena %TG: do we need a (\textit{Crocuta crocuta}) here?
%could resolve a 2 metre target at 1 km distance. 
While considering prehistoric habitats \cite{ruxton2004obligate} calculated that ``a 1 tonne mammal or reptile, in an ecosystem yielding carrion at densities similar to the current Serengeti, could have met its energy requirements if it could detect carrion over a distance of the order of 400–500 m".
The senses of many extant (and in all probabilty extinct) carnivores meet this required distance, making scavenging feasbile for terrstrial species \citep{farlow1994speculations,mech2010wolves}.

@@

Aside from sight, many birds have well developed olfactory systems \citep{AR:AR22815} including three species of vultures within the new world family Cathartidae, (genus \textit{Cathartes}).
% three species of vultures within the new world family Cathartidae, (genus \textit{Cathartes}), have well developed olfactory systems \citep{AR:AR22815}.
Among them are the Turkey Vultures (\textit{Cathartes aura}) which were able to locate 90\% of baits set out in a tropical forest \citep{houston1986olfaction}. An atuned sense of smell is obviously useful in detecting decaying carrion from the air.
% This would be impossible for the visually reliant old world species.

@@

% Depending on the species, a carcass in water either floats or descends to the sea floor \citep{Whitehead415}.
Aquatic species have to contend with the low-light environment of water where visual detection distances are far lower (< 100 m) than they would be in the air.
As such, animals here detect resources through chemo- and mechanoreception more so than through vision \citep{ruxton2004energetic}.
This is particularly relevant to extant sharks and aquatic snakes who are deemed as having the most suitable physiology for scavenging.
A hypothesis put forth by \cite{sazima1990necrofagia} argued that chemical gradients in water would allow for a relatively easier detection of carrion by snakes.
This gained some support from \cite{devault2002scavenging}, who found a preponderence of aquatic snake species in their review of this behaviour.
Smell seems to be the primary means of carcass detection in sharks as well. 
\cite{fallows2013white} found that wind speed determined the number of sharks feeding at whale carcasses indicating they were dependent on detecting the odours from the decaying whales.

@@

Interestingly, such comparisons have given insight into the feeding ecology of early hominins who, for instance, had the ability to craft tools for breaking open bones \citep{ARCM:ARCM12084}.
The question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene has been a matter of debate for years.
A recent study investigating potential scavenging opportunities for hominins in Kenya found that, even when discounting bone material, there is a substantial amount of scavengeable meat left on predated remains; sufficient to sustain the requirements of an adult male \textit{Homo erectus} \citep{pobiner2015new}.
In some historical hominin-inhabited areas there were a greater number of felids than hyenids.
Again, this is significant because hyenas are likely to have left far less flesh on a carcass than a felid such as a sabretooth enabling contemperaneous hominins to benefit \citep{pobiner2015new}.
The intelligence, resultant tool-use and cooperative nature of hominins meant they could likely adapt to take on more or less carrion depending on their environment \citep{moleon2014humans}.
%TG: or "could"? This is hypothetical behaviour stuff! Maybe past hyenas where exclusive sushi eaters!
%TG: Not sure about the intelligence thing: the sentence as the same meaning without and will be less anthropocentric sounding.

@@

Scavenging should be particularly attractive to avian predators compared to mammals. % AK should this section go above to where we mention avian scavengers first? 
Solitary mammalian predators can kill prey up to the same body mass as themselves and sometimes an order of magnitude heavier (e.g. socially hunting lions \citep{owen2008predator}).
In contrast, birds of prey tend to kill prey smaller than themselves \citep{slagsvold2007prey}.
This is likely due to their need to kill animals that they can fly away with, as well as the risk of injury being higher (which carries a higher mortality risk) for a bird than a mammal.
Scavenging provides a means for birds to exploit species that would otherwise be too big for them to kill.
% Thus, for birds, scavenging means they can exploit species that would otherwise be too big for them to kill.

@@

Perhaps the greatest environmental driver of scavenging tendency is that of temperature. 
The geological record shows the Earth has undergone radical fluctuations in temperature over time.
This will have had a significant bearing on the availability and persistence of carrion.
To illustrate the point, a 10$^{\circ}$C increase in ambient temperature can double carcass decomposition rates \citep{parmenter2009carrion} and geological evidence indicates that the Mesozoic Earth was at least 6 $^{\circ}$C warmer than now \citep{sellwood2006mesozoic}.
In terms of specific habitats, it has been shown that decomposition is greater in warm and moist areas versus more xeric ones \citep{beasley2015vertebrates}.
Moreover, oceanic productivity and habitat structure are all impacted by climactic conditions.
The impacts these can have on scavengers have been empirically supported e.g.
\cite{beasley2015vertebrates} who point to a series of studies showing how microbes and invertebrates benefit at higher temperatures to the detriment of vertebrate scavengers such that ``above 20$^{\circ}$C vertebrates were able to detect and consume only 19\% of small-mammal carcasses, whereas at temperatures below 18$^{\circ}$C, vertebrates consumed 49\% of carcasses".

%\section*{The Role of Climate} %TG: Use it as a opening at the end of the paper (i.e. conclusion + finishing with this bit)
\section*{Conclusion} 
As is often the case in science, the present provides the key to the past. The animals of today, while often different (sometimes radically so) to their ancestors, allow us to make informed comparisons to extinct species. 
We have used this technique to give insight into the drivers of scavenging across terrestrial vertebrates through time. In common with any other forager be they grazer, browser or predator, scavengers past and present have had to balance their energetic costs with the gains of food. 
Our four main axes: locomotion, detection, processing and environment can be used to create a scale of scavenging whereupon any species can be placed in order to establish the importance of carrion in it diet.  
% Add bits on climate change/dectection rate change, etc...
